Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-11 output/paper/paper.tex

@@

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv \and
	Jonas Wallin
}

@@

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' or ``capture-retain'' models (\cite{Pollock1991_review_papers}).

@@

\section{Model Performance}
To say judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by adding the mean historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations as confidence intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix). The central plot shows the width of the confidence intervals, and the rightmost one the proportion of CI´s that cover the true value.

@@

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes.

>>>>

\usepackage{amsmath}
\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}

@@

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Section of Sustainable Health, Umeå University}\thanks{Heidelberg Institute of Global Health, University of Heidelberg} \and
	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
}

@@

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' (\cite{Pollock1991_review_papers}).

@@

\section{Model Performance}
To say judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix). The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.

@@

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes. However, wew have noted a clear decline in proportions of deaths reported the two first working days. In early april .\% of deaths where reported within the first two working days wheras in late may only \% where reported during the same time interval. We adress this by assuming that the two deaths from the two first working days comes from a different population then the remainder of days.
